---
title: "Milestone Report"
author: "Tom Lous"
date: "25 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading & Summaries

For the purpose of creating a predictive input model, we have been supplied with a dataset [https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip), containing twitter, news and blog data for 4 languages.

For this part we'll be solely focussing on the English corpus.

```{r, cache=TRUE, warning=FALSE}
enTwitterLines <- readLines("data/final/en_US/en_US.twitter.txt")
enBlogLines <- readLines("data/final/en_US/en_US.blogs.txt")
enNewsLines <- readLines("data/final/en_US/en_US.news.txt")

wordCount <- function(lns){
  sum(sapply(gregexpr("\\S+", lns), length))
}

meanSentenceLength <- function(lns){
  mean(sapply(gregexpr("\\S+", lns), length))
}

lineCount <- function(lns){
  length(lns)
}

tRow <- c(lineCount(enTwitterLines), wordCount(enTwitterLines), meanSentenceLength(enTwitterLines))
bRow <- c(lineCount(enBlogLines), wordCount(enBlogLines), meanSentenceLength(enBlogLines))
nRow <- c(lineCount(enNewsLines), wordCount(enNewsLines), meanSentenceLength(enNewsLines))

infoEn <- rbind(tRow, bRow, nRow)
rownames(infoEn) <- c("twitter", "blog", "news")
colnames(infoEn) <- c("lines", "words", "mean.words.per.line")

infoEn
```

As we can see all datasets differ in amount of lines, but in total each has roughly between 30-37M words.


## Sampling

Since the English language has between 2k and 5k most common used words (depending on which source you lookup), we probably won't need all 100M words to be analyzed. 

There are ways to calculate the needed sample size, but they don't reaaly seem to apply to this problem, however.

If we just take 5% of all datasets we still end up with 5M words to be analyzed

*Check the utils package to see how sampling is done randomly*

```{r, cache=TRUE, warning=FALSE, results='hide'}
source("utils.R")
set.seed(11081979)

recreate <- FALSE
sampleFactor <- 0.05
info <- createAllSamples("data/final", sampleFactor, recreate)

twitterENInfo <- sampleFile("data/final/en_US/en_US.twitter.txt")
newsENInfo <- sampleFile("data/final/en_US/en_US.news.txt")
blogsENInfo <- sampleFile("data/final/en_US/en_US.blogs.txt")
```


These info objects contain the sample data & some meta information for later purposes
```{r}
str(twitterENInfo)
```

## Preprocessing

### 1. Tokenization
To actually parse the word's we are going to use a **whitespace tokenizer**.
First we'll split each line into sentences, because the end of a sentence should probably not be a predictor for the next one. This way we'll keep tokization independant of context.
Then we'll remove all non word chars, lowercase and split on whitespace. 
In the end we'll have a vector of term vectors

```{r, eval=FALSE}
tokenize <- function(dataset){
  dataset <- unlist(strsplit(dataset, "[\\.\\,!\\?\\:]+"))
  dataset <- tolower(dataset)
  dataset <- gsub("[^a-z\\s]", " ", dataset)
  dataset <- gsub("\\s+", " ", dataset)
  dataset <- trimws(dataset)
  dataset <- strsplit(dataset, "\\s")
  return(dataset)
}
```

```{r, echo=FALSE, warning=FALSE, results='hide'}
source("utils.R")
```

```{r, cache=TRUE}
twitterSampleENTokenized <- tokenize(twitterENInfo$sample.data)
newsSampleENTokenized <- tokenize(newsENInfo$sample.data)
blogSampleENTokenized <- tokenize(blogsENInfo$sample.data)
```

### 2. Merging

Since in essence the data is now uniformly for the EN corpus, we should combine these into one dataset

```{r, cache=TRUE}
sampleENTokenized <- c(twitterSampleENTokenized, newsSampleENTokenized, blogSampleENTokenized)
```

### 3. Filtering
Next remove some profanity we don't want in our dataset (we should never suggest profanity)

For each language we've downloaded a publicly kept profanity list from [https://github.com/LDNOOBW/List-of-Dirty-Naughty-Obscene-and-Otherwise-Bad-Words](https://github.com/LDNOOBW/List-of-Dirty-Naughty-Obscene-and-Otherwise-Bad-Words)

```{r, eval=FALSE}
profanityFilter <- function(termList, locale){
  profanities <- readLines(paste0("data/config/",locale,"/profanity.txt"))
  lapply(termList, setdiff, y=profanities)
}
```

```{r, echo=FALSE, warning=FALSE, results='hide'}
source("utils.R")
```

```{r, cache=TRUE}
sampleENTokenized <- profanityFilter(sampleENTokenized, twitterENInfo$locale)
```

We could use to filter other terms, like stopwords or 1 or 2 letter words, but although they probably will not be good predictors themselves (like: the, and, or) they will probably follow some other terms and could be predicted. Stripping them out will alter the syntacticle structure to much and therefore having a major effect on the predictions.


## Explore

### 1. View data
Now we have a sample tokenize dataset.
Let's explore

```{r}
head(sampleENTokenized, 3)

#Num lines
length(sampleENTokenized)

#Num terms
sum(sapply(sampleENTokenized, length))
```

We ended up with almost 5M terms divided over >700k sentences (term vectors)

### 2. Term frequencies

Using the frequencyTable function we can see what are the most common words in our dataset

```{r, eval=FALSE}
frequencyTable <- function(termList){
  term <- data.frame(unlist(termList))
  grouped <- as.data.frame(table(term))
  freq <- grouped[order(-grouped$Freq),]
  rownames(freq) <- 1:nrow(freq)
  
  total <- sum(freq$Freq)
  freq$CumFreq <- cumsum(freq$Freq)
  freq$Coverage <- freq$CumFreq/total
  
  return(freq)
}
```

```{r, echo=FALSE, warning=FALSE, results='hide'}
source("utils.R")
```


```{r, cache=TRUE}
sampleENTermFrequency <- frequencyTable(sampleENTokenized)

head(sampleENTermFrequency, 15)
```

```{r,fig.width=7, fig.height=6, echo=FALSE}
library(ggplot2)
tmp <- sampleENTermFrequency[1:50,]

tmp$termLength <-  nchar(as.character(tmp$term))

ggplot(tmp, aes(x=reorder(term,Freq), y=Freq, fill=termLength)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          panel.background = element_blank(),
          axis.title.y=element_blank(),
          axis.title.x=element_blank())
```
```{r,fig.width=7, fig.height=6, echo=FALSE}
library(ggplot2)
tmp <- filterFrequencyTable(sampleENTermFrequency, 0.005)


ggplot(tmp, aes(y=as.integer(rownames(tmp)), x=Coverage)) +
    geom_line() +
    coord_flip() + 
    labs(x="Coverage",y="Observations") +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          panel.background = element_blank()
          )
```
### 3. Check sample

These top terms seem a logical set.

To make sure we can compare our frequencies with a general dataset that collects this information, to see how examplary our sample is for the real word.
We downloaded the dataset [http://www.wordfrequency.info/top5000.asp](http://www.wordfrequency.info/top5000.asp)


```{r, cache=TRUE}
realENWordFrequency <- read.csv("data/config/en_US/word-frequency.csv")
realENWords <- tolower(as.character(unname(realENWordFrequency$Word)))

topLimit <- length(realENWords)

sampleENTermsSlice <- as.character(sampleENTermFrequency$term)[1:topLimit]

#Number of same top frequency words in
numIntersect <- length(intersect(realENWords,sampleENTermsSlice))
numIntersect

# Coverage factor
numIntersect/topLimit 
```

A coverage factor of 0.57 is maybe not the best, but good enought to continu.
Perhaps increasing the samplesize will retun better results, or validate the top list

### 4. Create n-grams

Now we have term vectors and we are reasonably sure about the quality, we can now create n-grams using this function:

```{r, eval=FALSE}
createNgram <- function(vec, n=2){
  l <- length(vec) 
  if(l < n){
    return(c())
  }else if(l == n){
    return(paste(vec, collapse = " "))
  }else{
    numNgrams <- l-n+1
    mtrx <- matrix(nrow=numNgrams, ncol=n)
    for(i in 1:n){
      m <- l - n + i
      mtrx[,i] <- vec[i:m]
    }
    ngrams <- apply(mtrx, 1, paste, collapse=" ")
    return(ngrams)
  }
} 

transformNGram <- function(termList, n=2){
  lapply(termList, createNgram, n=n)
}
```


### 5. Bi-gram

```{r, cache=TRUE}
sampleENBiGrams <- transformNGram(sampleENTokenized, 2)

sampleENBiGramsFrequency <- frequencyTable(sampleENBiGrams)

head(sampleENBiGramsFrequency, 15)
```

```{r,fig.width=7, fig.height=6, echo=FALSE}
library(ggplot2)
tmp <- filterFrequencyTable(sampleENBiGramsFrequency, 0.005)


ggplot(tmp, aes(y=as.integer(rownames(tmp)), x=Coverage)) +
    geom_line() +
    coord_flip() + 
    labs(x="Coverage",y="Observations") +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          panel.background = element_blank()
          )
```

### 6. Tri-gram

```{r, cache=TRUE}
sampleENTriGrams <- transformNGram(sampleENTokenized, 3)

sampleENTriGramsFrequency <- frequencyTable(sampleENTriGrams)

head(sampleENTriGramsFrequency, 15)
```
```{r,fig.width=7, fig.height=6, echo=FALSE}
library(ggplot2)
tmp <- filterFrequencyTable(sampleENTriGramsFrequency, 0.005)


ggplot(tmp, aes(y=as.integer(rownames(tmp)), x=Coverage)) +
    geom_line() +
    coord_flip() + 
    labs(x="Coverage",y="Observations") +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          panel.background = element_blank()
          )
```


### 7. Coverage overview

```{r, eval=FALSE}
coverageFactor <- function(freqTable, coverage){
  pos <- nrow(freqTable[freqTable$Coverage < coverage,])
  pos / nrow(freqTable) 
}
```

Let's see what the coverage of the n-gram sets is compare to the entire corpus

```{r}
coverageFactors <- c(0.1,0.5,0.9)
uniCov <- sapply(coverageFactors, coverageFactor, freqTable=sampleENTermFrequency)
biCov <- sapply(coverageFactors, coverageFactor, freqTable=sampleENBiGramsFrequency)
triCov <- sapply(coverageFactors, coverageFactor, freqTable=sampleENTriGramsFrequency)

infoCov <- rbind(uniCov, biCov, triCov)
rownames(infoCov) <- c("uni-gram", "bi-gram", "tri-gram")
colnames(infoCov) <- coverageFactors

infoCov
```



# Evaluation

## 1. Prediction

We see that small parts of the data are responsible for the bulk of the corpus. This allows prediction to be a smaller model to just focus on the most important parts.


## 2. Next steps

* Reevaluate approach and see if sample size adjust,inclusion of stopwords, punctuation, numbers, etc improve prediction
* Building a predictive model using the identified tokens
* Wrapping up the results and the developed model as a data product, shiny app.
